TacticToe: Learning to Prove with Tactics

### 2018-09-03 tactictoe_jv.tex
minor tweaks

\paragraph{Term}
Some proofs require to instantiate a tactic argument by a term. Here are two 
abstracted 
tactics that use term 
arguments with placeholders \texttt{term} and \texttt{thm}: 
  \[\texttt{EXISTS\_TAC term},\ \texttt{SPEC\_THEN term STRUCT\_CASES\_TAC 
  thm}\]
Finding a witness to instantiate an existential formula is challenging because 
it is likely that this witness may not have been proposed before and there is 
no 
general rule on how to generate it. Instantiating 
the term argument in the second tactic is easier because the expected term is a 
variable of the goal. That is why we think that the first step 
toward term argument prediction is to select subterms of the goal ordered by 
their similarity with the term argument appearing in the original tactic. 
Additional experiments, using a tactic that tries the 16 closest subterms 
$[term1,\ldots,term16]$ to the goal with the trial and error procedure 
described in the previous paragraph, show that this simple term predictor 
improves the success 
rate of 
\tactictoe by another four percent. With the modified 
\texttt{ORELSE} tactical, the tactic derived from the abstraction
\texttt{EXISTS\_TAC term} in this experiment can be written as:
   \[\texttt{EXISTS\_TAC term1 ORELSE ... ORELSE EXISTS\_TAC term16}\]

@@

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} can
generate
proof scripts from previous \coq proof examples.
Its strength lies in the ability to produce likely sequences
of tactics for solving domain specific goals. It operates by creating a model
for common sequences of tactics for a specific library.
This means that in order to propose the next tactic, only the previously
called tactics
are considered.
Our algorithm, on the other hand, relies mainly on the characteristics of the
current goal
to decide
which tactics to apply next. In this way, our learning mechanism has to
rediscover why each
tactic was applied for the current subgoals. It may lack some useful bias for
common sequences
of tactics, but is more reactive to subtle changes. Indeed, it can be trained
on a large library and only tactics relevant to the current subgoal will be
selected.
Concerning the proof search, \textsf{SEPIA}'s %simple
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the
\coq library demonstrating that it globally outperforms individual \coq
tactics. Here, we demonstrate the competitiveness of our system against
\eprover on the \holfour standard library.

@@

A closely related approach is proof planning that was first described in 
~\cite{DBLP:conf/cade/Bundy88}. Proof planners have been implemented to 
facilitate interactive theorem proving 
development in Oyster~\cite{DBLP:conf/cade/BundyHHS90},  
Omega~\cite{DBLP:conf/cade/BenzmullerCFFHKKKMMSSS97} and 
\isabelle~\cite{DBLP:conf/cade/DixonF03}.
A proof plan can be seen as a program that given a goal 
returns the appropriate tactic for this goal. From this point of view, 
\tactictoe is a general proof plan that learns to apply the right tactic from 
examples and possess a backtracking mechanism given by the MCTS algorithm.
The philosophy of proof planning is also to create a few 
tactics that can prove a wide range of theorems. % In this respect,  if we omit 
Apart from the 
abstraction technique, \tactictoe is not yet synthesizing any tactics but simply 
selecting them out of a large tactic set. This may become necessary in the future, 
using a variety of statistical/symbolic learning methods.
% In the end, proof planning methodology 
% may become necessary for improving \tactictoe as it provides a way to generate 
% tactic applicable to a 
% large range of formulas. 
More general tactics are usually easier to predict 
accurately by machine learning models as they are associated with more 
(successful) examples in our database.

@@

Machine learning has also been used to advise the best library lemmas for new
ITP goals.
This can be done either in an interactive way, when the user completes the
proof based on the recommended lemmas, as in the Mizar proof advisor
\textsc{MizAR}~\cite{Urb04-MPTP0,mizAR40}, or attempted fully automatically, where such lemma
selection is handed over to the ATP component of a \emph{hammer}
system, as in the hammers for \holfour~\cite{tgck-cpp15}, 
\hollight~\cite{holyhammer} and \isabelle~\cite{BlanchetteGKKU16}.

>>>>

\paragraph{Term}
Some proofs require to instantiate a tactic argument by a term. Here are two 
abstracted 
tactics that use term 
arguments with placeholders \texttt{term} and \texttt{thm}: 
  \[\texttt{EXISTS\_TAC term},\ \texttt{SPEC\_THEN term STRUCT\_CASES\_TAC 
  thm}\]
Finding a witness to instantiate an existential formula is challenging because 
it is likely that this witness may not have been proposed before and there is 
no 
general rule to generate it. Instantiating 
the term argument in the second tactic is easier because the expected term is a 
variable of the goal. That is why we think that the first step 
toward term argument prediction is to select subterms of the goal ordered by 
their similarity with the term argument appearing in the original tactic. 
Additional experiments, using a tactic that tries the 16 closest subterms 
$[term1,\ldots,term16]$ to the goal with the trial and error procedure 
described in the previous paragraph, show that this simple term predictor 
improves the success 
rate of 
\tactictoe by another four percent. With the modified 
\texttt{ORELSE} tactical, the tactic derived from the abstraction
\texttt{EXISTS\_TAC term} in this experiment can be written as:
   \[\texttt{EXISTS\_TAC term1 ORELSE ... ORELSE EXISTS\_TAC term16}\]

@@

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} can
generate
proof scripts from previous \coq proof examples.
Its strength lies in the ability to produce probable sequences
of tactics for solving domain specific goals. It operates by creating a model
for common sequences of tactics for a specific library.
This means that in order to propose the next tactic, only the previously
called tactics
are considered.
Our algorithm, on the other hand, relies mainly on the characteristics of the
current goal
to decide
which tactics to apply next. In this way, our learning mechanism has to
rediscover why each
tactic was applied for the current subgoals. It may lack some useful bias for
common sequences
of tactics, but is more reactive to subtle changes. Indeed, it can be trained
on a large library and only tactics relevant to the current subgoal will be
selected.
Concerning the proof search, \textsf{SEPIA}'s %simple
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the
\coq library demonstrating that it globally outperforms individual \coq
tactics. Here, we demonstrate the competitiveness of our system against
\eprover on the \holfour standard library.

@@

A closely related approach is proof planning that was first described in 
~\cite{DBLP:conf/cade/Bundy88}. Proof planners have been implemented to 
facilitate interactive theorem proving 
development in Oyster~\cite{DBLP:conf/cade/BundyHHS90},  
Omega~\cite{DBLP:conf/cade/BenzmullerCFFHKKKMMSSS97} and 
\isabelle~\cite{DBLP:conf/cade/DixonF03}.
A proof plan can be seen as a program that given a goal 
returns the appropriate tactic for this goal. From this point of view, 
\tactictoe is a general proof plan that learns to apply the right tactic from 
examples and has a backtracking mechanism given by the MCTS algorithm.
The philosophy of proof planning is also to create a few 
tactics that can prove a wide range of theorems. % In this respect,  if we omit 
Apart from the 
abstraction technique, \tactictoe is not yet synthesizing any tactics but simply 
selecting them out of a large tactic set. This may become necessary in the future, 
using a variety of statistical/symbolic learning methods.
% In the end, proof planning methodology 
% may become necessary for improving \tactictoe as it provides a way to generate 
% tactic applicable to a 
% large range of formulas. 
More general tactics are usually easier to predict 
accurately by machine learning models as they are associated with more 
(successful) examples in our database.

@@

Machine learning has also been used to advise the best library lemmas for new
ITP goals.
This can be done either in an interactive way, when the user completes the
proof based on the recommended lemmas, as in the Mizar Proof Advisor~\cite{Urb04-MPTP0}, or attempted fully automatically, where such lemma
selection is handed over to the ATP component of a \emph{hammer}
system, as in the hammers for \holfour~\cite{tgck-cpp15}, 
\hollight~\cite{holyhammer}, \isabelle~\cite{BlanchetteGKKU16} and \mizar~\cite{UrbanRS13,mizAR40}.
